---
title: "Smartphone User Activities"
author: "Aleksej Hoffärber"
date: "10/1/2020"
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

source("packages.R")
options(width = 100, digits = 2)
theme_set(theme_minimal())
```

## Goals and Procedure

The aim of this short project was to get familiar with the DNN possibilities in R, especially while using caret, RSNSS, and h2o. Important emphasis was put on regularization, visualization, and variable importance techniques in order to make the results easily comprehensible. In terms of models, only basic DNN architectures were employed with a varying number and values of hyperparameters in order to find optimal models. 

The main dataset to which most of the code and plots refer displays sensor data of smartphone users in order to predict their behavior. Other plots refer to a movie data set in order to visualize residual predictions.

## Plot Overview

1. Hyperparameter Tuning 
2. GAM-based Hyperparameter Selection
3. Variable Importance Analysis
4. Residual Prediction Analysis (movie dataset)

```{r hand-writing classification, echo=F, warning=F, results=F, message=F}
d_train <- read.csv("data/train.csv") %>% 
  mutate(label = as.factor(label))


c2 <- h2o.init(max_mem_size = "20G", nthreads = 10)
h2od <- as.h2o(d_train, # data upload
                    destination_frame = "h2odigits")

# training and test set
i <- 1:20000
h2od_train <- h2od[i,-1]

it <- 20001:30000
h2od_test <- h2od[it,-1]
xnames <- colnames(h2od_train)

# model tuning

if (exists("m3") == FALSE) 
{m3 <- h2o.deeplearning(
  x = xnames,
  training_frame = h2od_train,
  validation_frame = h2od_test,
  activation = "TanhWithDropout",
  autoencoder = T,
  hidden = c(100,10),
  epochs = 30,
  sparsity_beta = 0,
  input_dropout_ratio = 0,
  hidden_dropout_ratios = c(0,0),
  l1 = 0,
  l2 = 0
)} 



error3 <- as.data.frame(h2o.anomaly(m3, h2od_train))
percentile3 <- quantile(error3$Reconstruction.MSE, probs = .99)

# CREATE DF WITH DL RESULTS

f3 <- as.data.frame(h2o.deepfeatures(m3, h2od_train, 2)) # 2nd layer
f3$label <- d_train$label[i]


colnames(f3) <- c(rep(1:10), "label")


f3 %>% group_by(label) %>% 
  summarise_all(mean) %>% 
  pivot_longer(c(2:11)) %>% # long format for plotting
  as.data.frame() %>% 
  mutate(name = I(as.integer(name) -1)) -> f3c 
  # important to keep name otherwise no line possible 


layer <- 
  f3c %>% 
  ggplot(aes(x = name, y = value, 
             col = label, linetype = label)) + # line per label
  geom_line(lwd = 1) +
  scale_x_continuous("Deep Features eq. to 2nd Hidden Layer", 
                     breaks = c(1:10), n.breaks = 10) +
  labs(title = "Probability Assignment",
       subtitle = "Despite structure, 2nd layer did not result in a clear classification seperation") +
  theme_minimal() +
  theme(legend.position = "bottom", legend.key.width = unit(1, "cm"))

print(layer)
ggsave("readme_files/figure-markdown_github/hand-writing-classification-1.png")



```

```{r user activities, include=FALSE, echo=F, warning=F, results=F, message=F}
# rm(list = ls())

# data
us_train.x <- read.table("data/UCI HAR Dataset/train/X_train.txt")
us_train.y <- read.table("data/UCI HAR Dataset/train/y_train.txt")[[1]]

us_test.x <- read.table("data/UCI HAR Dataset/test/X_test.txt")
us_test.y <- read.table("data/UCI HAR Dataset/test/y_test.txt")[[1]]

us_train <- cbind(us_train.x, Outcome = factor(us_train.y))
us_test <- cbind(us_test.y, Outcome = factor(us_test.y))
us_labels <- read.table("data/UCI HAR Dataset/activity_labels.txt")

 # cluster
c1 <- h2o.init(max_mem_size = "30G",
               nthreads = 7)

# random serach & performance check
run <- function(seed, 
                name = paste0("m_", seed), 
                run = T) 
  {
  
  set.seed(seed)

  p <- list(
    Name = name,
    seed = seed,
    depth = sample(1:5,1),
    
    l1 = runif(1, 0, .01),
    l2 = runif(1, 0, .01),
    
    input_dropout = rbeta(1, 1, 12),
    rho = runif(1, .9, .999),
    epsilon = runif(1, 1e-10, 1e-8)
  )

  p$neurons <- sample(20:400, p$depth, replace =T)
  p$hidden_dropout <- rbeta(p$depth, 1.5, 1)/2

  if (run) {
    model <- h2o.deeplearning(
      x = colnames(us_train.x),
      y = "Outcome",
      training_frame = h2o.tr,
      activation = "RectifierWithDropout",
      hidden = p$neurons,
      epochs = 100,
      loss = "CrossEntropy",
      input_dropout_ratio = p$input_dropout,
      hidden_dropout_ratios = p$hidden_dropout,
      l1 = p$l1,
      l2 = p$l2,
      rho = p$rho,
      epsilon = p$epsilon,
      
      export_weights_and_biases = TRUE,

      model_id = p$Name
    )
  
    ## performance on training data
    p$MSE <- h2o.mse(model)
    p$R2 <- h2o.r2(model)
    p$Logloss <- h2o.logloss(model)
    p$CM <- h2o.confusionMatrix(model)
    ## performance on testing data
    perf <- h2o.performance(model, h2o.te)
    p$T.MSE <- h2o.mse(perf)
    p$T.R2 <- h2o.r2(perf)
    p$T.Logloss <- h2o.logloss(perf)
    p$T.CM <- h2o.confusionMatrix(perf)
  
    } 
    else 
      {
       model <- NULL
  
      }
    return(list(
      Params = p,
      Model = model))
}

h2o.tr <- as.h2o(
  us_train,
  destination_frame = "h2o.tr")

h2o.te <- as.h2o(
  us_test,
  destination_frame = "h2o.te")

# specifiying seeds
us.seeds <- c(403L, 10L, 329737957L, -753102721L, 1148078598L, -1945176688L,
               -1395587021L, -1662228527L, 367521152L, 217718878L, 1370247081L,
               571790939L, -2065569174L, 1584125708L, 1987682639L, 818264581L,
               1748945084L, 264331666L, 1408989837L, 2010310855L, 1080941998L,
               1107560456L, -1697965045L, 1540094185L, 1807685560L, 2015326310L,
               -1685044991L, 1348376467L, -1013192638L, -757809164L, 1815878135L,
               -1183855123L, -91578748L, -1942404950L, -846262763L, -497569105L,
               -1489909578L, 1992656608L, -778110429L, -313088703L, -758818768L,
               -696909234L, 673359545L, 1084007115L, -1140731014L, -877493636L,
               -1319881025L, 3030933L, -154241108L, -1831664254L)

if (exists("model.res") == FALSE) 
  {model.res <- lapply(us.seeds, run)} 


# validation
model.res.dat <- do.call(rbind, lapply(model.res, function(x)
with(x$Params,
     data.frame(l1 = l1, l2 = l2,
                depth = depth, input_dropout = input_dropout,
                SumNeurons = sum(neurons),
                MeanHiddenDropout = mean(hidden_dropout),
                rho = rho, 
                epsilon = epsilon, 
                MSE = T.MSE))))

p.perf <- ggplot(melt(model.res.dat, id.vars = c("MSE")), aes(value, MSE)) +
  geom_point() +
  stat_smooth(color = "black") +
  theme(text = element_text(size = 8))+
  facet_wrap(~ variable, scales = "free_x", ncol = 2)

print(p.perf)
ggsave("readme_files/figure-markdown_github/hyperparameter-selection.png")


```

```{r GAM training, echo=FALSE, warning=F, results=F, message=F}

# GAM training
m.gam <- gam(MSE ~ s(l1, k = 4) +
               s(l2, k = 4) +
               s(input_dropout) +
               s(rho, k = 4) +
               s(epsilon, k = 4) +
               s(MeanHiddenDropout, k = 4) +
               te(depth, SumNeurons, k = 4),
             data = model.res.dat)


png(filename = "readme_files/figure-markdown_github/gam-hyperparameter-selection.png",
    width = 1000, height = 880)

par(mfrow = c(3, 2))

for (i in 1:6) {
  plot(m.gam, select = i)
}

dev.off()

png(filename = "readme_files/figure-markdown_github/hyperparameter-interaction.png",
    width = 1000, height = 880)

par(mfrow = c(1,1))
plot(m.gam, select = 7)
dev.off()
```

```{r optimized model, warning=F, results=F, message=F}

if (exists("model.optimized") == FALSE) {
  model.optimized <- h2o.deeplearning(
    x = colnames(us_train.x),
    y = "Outcome",
    training_frame = h2o.tr,
    validation_frame = h2o.te,
    activation = "RectifierWithDropout",
    hidden = c(300, 300, 300),
    epochs = 100,
    loss = "CrossEntropy",
    input_dropout_ratio = .08,
    hidden_dropout_ratios = c(.50, .50, .50),
    l1 = .002,
    l2 = 0,
    rho = .95,
    epsilon = 1e-10,
    
    diagnostics = TRUE,
    export_weights_and_biases = TRUE,
    variable_importances = TRUE,
    model_id = "optimized_model"
)
}
```

```{r variable importance optimized model, echo=FALSE, warning=F, results=F, message=F}

h2o.performance(model.optimized)

model.res.dat[which.min(model.res.dat$MSE), ]

imp2 <- as.data.frame(h2o.varimp(model.optimized))

pr.imp <- imp2 %>% arrange(desc(percentage)) %>% slice(1:50) %>% 
  ggplot(aes(x = factor(variable, levels = variable), y = percentage)) +
  geom_point(color = "cornflowerblue") +
  geom_hline(yintercept=0.01) +
  geom_hline(yintercept=0.02) +
  theme_minimal() +
  theme(axis.text.x = element_text(
    angle = 90, vjust = 1, hjust = 1, size = 5)) +
  labs(title = "Variable Importance",
       subtitle = "Clear importance distinction for first 17 predictors") +
  xlab("Variables") +
  ylab("Score")

print(pr.imp)
ggsave("readme_files/figure-markdown_github/variable-importance-users.png")


```


```{r movies, echo=FALSE, warning=F, results=F, message=F}
c1 <- h2o.init(max_mem_size = "14G",
               nthreads = 6)


# NEW DATA:
# unzip("YearPredictionMSD.txt.zip")

d <- fread("YearPredictionMSD.txt", sep = ",") # speedup

ggplot(d[,.(V1)], aes(V1)) +
  geom_histogram(binwidth = 1) +
  xlab("Year of Release")

d %>% 
  group_by(V1) %>% 
  add_count() %>% 
  summarise(mean = mean(n)) -> da

# Cleaning for "out of quantile" observations
d %<>% filter(V1 >= quantile(d$V1, probs = c(0.005, 0.995))[1] &
                V1 <= quantile(d$V1, probs = c(0.005, 0.995))[2]) 


# Test/train
split <- 0.9

dtr <- d %>% slice(1:(as.integer(nrow(d)) * split))
dte <- d %>% slice((as.integer(nrow(d) * split + 1)):(as.integer(nrow(d) + 1)))

# h2o clustering
h2o.tr <- as.h2o(dtr, destination_frame = "h2o.tr")
h2o.te <- as.h2o(dte, destination_frame = "h2o.te")

# model fitting
summary(m0 <- lm(V1 ~., data = dtr)) # 24% in variance
cor(dte$V1,
    predict(m0, newdata = dte))^2 # 23% variance in test set

if (exists("m1") == FALSE) {
  m1 <- h2o.deeplearning(
    x = colnames(d)[-1],
    y = "V1",
    training_frame= h2o.tr,
    validation_frame = h2o.te, # for instant test performance output
    activation = "RectifierWithDropout",
    hidden = c(50),
    epochs = 100,
    input_dropout_ratio = 0,
    hidden_dropout_ratios = c(0),
    score_training_samples = 0,
    score_validation_samples = 0, # performance scoring on full data set
    diagnostics = TRUE,
    export_weights_and_biases = TRUE,
    variable_importances = TRUE
)
}

# h2o.saveModel(
#   object = m1,
#   path = paste0(getwd(),"\\m1"),
#   force = T)

# residuals
yhat <- as.data.frame(h2o.predict(m1, h2o.tr))
yhat <- cbind(as.data.frame(h2o.tr[["V1"]]), yhat)


p.resid <- ggplot(yhat, aes(factor(V1), predict - V1)) +
  geom_boxplot(fill = "cornflowerblue",
               outlier.colour = "darkred",
               outlier.size = 1) +
  geom_hline(yintercept = 0) +
  theme_minimal() +
  theme(axis.text.x = element_text(
    angle = 90, vjust = 1, hjust = 1)) +
  labs(title = "Prediction Residuals",
       subtitle = "Model results indicate bad prediction fits in the first years, because of skewed data") +
  xlab("Year of Release") +
  ylab("Difference in Predicted Years of Release")

print(p.resid)
ggsave("readme_files/figure-markdown_github/residual-predictions.png")


# variable importance
imp <- as.data.frame(h2o.varimp(m1))

p.imp <- imp %>% 
  ggplot(aes(x = factor(variable, levels = variable), y = percentage)) +
  geom_point(color = "cornflowerblue") +
  geom_hline(yintercept=0.01) +
  geom_hline(yintercept=0.02) +
  theme_minimal() +
  theme(axis.text.x = element_text(
    angle = 90, vjust = 1, hjust = 1, size = 5)) +
  labs(title = "Variable Importance",
       subtitle = "Clear importance distinction for first 11 predictors") +
  xlab("Variables") +
  ylab("Score")

print(p.imp)
ggsave("readme_files/figure-markdown_github/variable-importance-movies.png")


```

